A Practical Optional Type System for Clojure

### 2012-10-28 dissertation/design-choices.tex
Edit Java design section for clarity

Occurrence typing, described by Tobin-Hochstadt and Felleisen~\cite{TF10},
plays a major role in capturing many common Clojure idioms.
One important idiom it helps Typed Clojure to understand 
is the \emph{sequence abstraction}, which
is made up of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.

@@

This function type takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
constructs a union of types \lstinline|t|), and returns a type \lstinline|(U nil (Seqable x))|.
Occurrence typing allows us to attach propositions to the return
type of a function. Roughly, these say if \lstinline|(seq coll)|
returns a true value, then \lstinline|coll| is of type
\lstinline|(CountRange 1)| (a type representing a sequence of length 1 or greater),
otherwise \lstinline|coll| is \emph{either} of type \lstinline|nil| or of type \lstinline|(ExactCount 0)| (a type
representing an empty sequence).
(See section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).

@@

Strickland, Tobin-Hochstadt, and Felleisen describe variable-arity poly\-morphism ~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
as described in section \ref{sec:impvariablearity}

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland et al.\ .
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (section \ref{sec:futurearity}).

@@

% what is this paragraph saying?
Dialects of Clojure provide fast and unrestricted access to their host platform. 
This is a intentional source of incompatibility between Clojure implementations.
For example, Clojure is hosted on the Java Virtual Machine (JVM), ClojureCLR on the Common Language Runtime,
and ClojureScript on Javascript VMs. These are very different platforms, and it is unlikely
a Clojure dot call will be portable. There is no attempt at reconciling
host interoperability differences, and it is up to the programmer to decide
how to best abstract over different hosts.

Typed Clojure targets the JVM hosted Clojure.
Clojure embraces the JVM as a host by sharing its runtime type system and encouraging direct
interaction with libraries written in other JVM languages. Most commonly, Clojure programmers
reach to Java libraries. Java is statically typed language. It follows that any interaction with
Java will already have an annotated Java type. Typed Clojure can use these annotations
to statically type check these interactions.

Typed Clojure attempts to improve some shortcomings of Java's static type system.

Java's static type system does not provide a type-safe construct for eliminating
the \lstinline|null| pointer. Instead, Java programmers often rely on either testing
for \lstinline|null| or prior domain knowledge.

\begin{lstlisting}[caption=null elimination in Java]
...
Object a = nextObject();
if (a != null)
  parseNextObject(a);
else
  throw new Exception("null Pointer Found!);
...
\end{lstlisting}

In Java, \lstinline|null| is included in all reference
types, but \lstinline|null| is not expressible as a static type. In other words,
when a Java type claims a reference type, you should also assume it can also be of 
type \lstinline|null|.

Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
where \lstinline|null| might occur. This approach is not appropriate for Typed Clojure
as pattern matching is not idiomatic in Clojure.

@@

\subsection{Interaction with null}
\label{sec:interactionnull}

It is a goal of Typed Clojure to statically prevent many Java \emph{NullPointerException}
exceptions.
\emph{null} is explicitly expressible in Typed Clojure as the singleton type \lstinline|nil|,
representing the \emph{value} \lstinline|nil|.
This enables more expressive types that specify where \emph{nil} is and is not allowed.
Section \ref{sec:javainterop} gives an example of how this could be useful 
when using Java interoperability and gives details on the different ways Typed Clojure.

Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
treatment of \lstinline|nil|.
It allows us to track whether particular bindings are possibly \lstinline|nil|
down conditional branches, and makes it practical to throw type errors 
where \lstinline|nil| is not allowed (see section \ref{sec:OccurrenceTyping} for examples).
This is not practical in Java's type system because \emph{null} is not directly
expressible, which makes it impractical to error when a \emph{NullPointerException}
exception is possible: according to the type system, almost every operation
could involve \emph{null}!
Eliminating \emph{NullPointerException} exceptions is entirely the programmers
responsibility in Java.
Typed Clojure helps the programmer by statically proving certain misuses of \lstinline|nil| are impossible.

The design choice of potentially misusing \lstinline|nil| results in a type error
needs further work to be completely satisfying.
Certain cases of type inference should be improved to avoid redundant code
that ``pleases the type checker''.
For example, we currently do not know how to infer \lstinline|(filter identity coll)|
as returning a sequence of non-nil elements, for any sequence \lstinline|coll|.
(see section \ref{sec:filterneg}).
It is also possible for code interacting with Java to introduce unexpected \emph{null}
values. Currently, we let the programmer define a set of unenforced \emph{assumptions}
where \emph{null} is allowed in Java fields and methods (see section \ref{sec:javainterop}).
Adding runtime assertions that enforce these assumptions would allow more
accurate errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
interfacing with untyped Racket code.

@@

Common usages of datatypes are checkable, like datatype
constructors, field accesses, and protocol method invocation.
These are described in section \ref{sec:datatypes}.
Mutable fields in datatypes are not yet supported, but is
also a rarely used feature.

@@

Heterogeneous maps types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys with the type associated with each key in
a map tracked separately, similar to type systems for record types 
(see work by Harper and Pierce~\cite{HP91}).
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

>>>>

Occurrence typing, described by Tobin-Hochstadt and Felleisen~\cite{TF10},
plays a major role in capturing many common Clojure idioms.
One such idiom is the \emph{sequence abstraction}, which
consists of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.

@@

This function type takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
constructs a union of types \lstinline|t|), and returns a type \lstinline|(U nil (Seqable x))|.
Occurrence typing allows us to attach propositions to the return
type of a function. Roughly, these say if \lstinline|(seq coll)|
returns a true value, then \lstinline|coll| is of type
\lstinline|(CountRange 1)| (a type representing a sequence of length 1 or greater),
otherwise \lstinline|coll| is \emph{either} of type \lstinline|nil| or of type \lstinline|(ExactCount 0)| (a type
representing an empty sequence).
(See Section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).

@@

Strickland, Tobin-Hochstadt, and Felleisen describe variable-arity poly\-morphism ~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
as described in Section \ref{sec:impvariablearity}

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland et al.\ .
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (Section \ref{sec:futurearity}).

@@

Clojure implementations are designed to target popular platforms, 
providing fast, unrestricted access to their respective hosts.
Notable implementations target
the Java Virtual Machine (via Clojure), the Common Language Runtime (via ClojureCLR),
and Javascript virtual machines (via ClojureScript). 
Host interoperability is the main source of incompatibility between Clojure implementations.
Aside from syntactic matters, Clojure makes no attempt at reconciling
host interoperability differences:
it is up to the programmer to decide how to best write portable Clojure code.

Typed Clojure is designed for Clojure, which runs on the Java Virtual Machine (JVM).
Clojure embraces the JVM as a host by sharing its runtime type system and providing direct
interaction with libraries written in other JVM languages. Most commonly, Clojure programmers
use libraries written in Java. Java is a statically typed language, and it follows that any interaction with
Java will already have an annotated Java type. Typed Clojure uses Java's static type annotations
to help statically type check interactions with Java from Clojure.

This section details where Typed Clojure attempts to improve some shortcomings of Java's static type system.

@@

\subsection{Interaction with \emph{null}}
\label{sec:interactionnull}

As of Java 7, Java does not provide a type-safe construct for eliminating
occurrences of \emph{null}. Instead, Java programmers rely on ad-hoc approaches,
like testing for the presence of \emph{null} (like Listing \ref{lst:nullelim}) or prior domain knowledge.
One solution to this problem is an Option type (also known a Maybe type).
Scala provides \emph{scala.Option} for this purpose. The programmer provides two branches of code,
(respectively to execute if an expression is \emph{null}, and if it is not)
and the appropriate branch is chosen at runtime.
There is no such construct in idiomatic Clojure, so this solution is inappropriate:
a satisfactory solution should reuse current Clojure idioms.

\begin{lstlisting}[caption=null elimination in Java, label=lst:nullelim]
...
Object a = nextObject();
if (a != null)
  parseNextObject(a);
else
  throw new Exception("null Pointer Found!);
...
\end{lstlisting}

It is a goal of Typed Clojure to statically prevent misuses of Java's \emph{null}.
At runtime, Clojure represents \emph{null} as the value \lstinline|nil|;
Typed Clojure assigns this value the static type \lstinline|nil|
(a singleton type containing just the value \lstinline|nil|).
This enables more expressive types that explicitly communicate where \lstinline|nil|,
and therefore \emph{null}), is allowed.
Section \ref{sec:javainterop} gives an example of how this could be useful 
when using Java interoperability.

The crucial aspect of this approach is that misusing \emph{null} results in a \emph{static type error}.
This differs significantly from Java: according to Java's static type system,
it is implicit that almost every operation could involve \emph{null}!
Occurrence typing~\cite{TF10} plays a crucial role in making this a practical
design choice, by understanding existing Clojure idioms related to \emph{null} elimination (see Section \ref{sec:OccurrenceTyping}).
Eliminating occurrences of \emph{null} in Java is the programmer's responsibility.
Instead, Typed Clojure helps the programmer by statically proving many misuses of \emph{null} are impossible.

This decision, however, needs further work to be completely satisfying.
In particular, type inference should be improved in certain common situations.
For example, we currently do not know how to infer \lstinline|(filter identity coll)|
as returning a sequence of non-\lstinline|nil| elements, for any sequence \lstinline|coll|
(see Section \ref{sec:filterneg}).
Also, Typed Clojure lets the programmer define a set of unenforced \emph{assumptions}
where \emph{null} is allowed in Java fields and methods (see Section \ref{sec:javainterop}),
which is potentially brittle: there is nothing stopping the Java code from introducing unexpected \emph{null} values at runtime.
Adding runtime assertions that enforce these assumptions would allow more
accurate and earlier errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
interfacing with untyped Racket code.

@@

Common usages of datatypes are checkable, like datatype
constructors, field accesses, and protocol method invocation.
These are described in Section \ref{sec:datatypes}.
Mutable fields in datatypes are not yet supported, but is
also a rarely used feature.

@@

Heterogeneous maps types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (see Section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys with the type associated with each key in
a map tracked separately, similar to type systems for record types 
(see work by Harper and Pierce~\cite{HP91}).
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.
